---
title: "Introduction to the landmaRk package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the landmaRk package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```




# Overview

The landmaRk package provides a framework for landmarking analysis of
time-to-event data with time-varying covariates. It allows users to perform
survival analysis using longitudinal data, fitting models to the time-varying
covariates, and then using these predictions in survival models.

Given a time-to-event outcome $T_i$, a landmark time $s$ and a time horizon 
$s + w$, the goal of a landmarking analysis is to estimate 
$$
\pi_i(s+w \vert s) = P(T_i > s+w \vert T_i \ge s, X_i(s)), 
$$
where $X_i(s)$ denotes a vector of covariates which may include time-varying 
covariates.

A landmarking analysis of time-to-event data has two components:

- First, model the longitudinal trajectories of dynamic covariates, $X_i(t)$. 
Then use, the fitted model to make a prediction for $X_i(s)$, $\hat{X}_i(s)$, 
at the landmark time, $s$.

- Second, fit a survival model of the time-to-event outcome, conditioning on the
predicted value for $X_i(s)$, and potentially on additional static and dynamic 
covariates. 

The `landmaRk` package allows users to use the following method for the first 
component:

- Last Observation Carried Forward (LOCF), using the last measurement for $X_i$
recorded prior to $s$ as our prediction, $\hat{X}_i(s)$.

- Linear mixed-effect (LME) model, as implemented in the `lme4` package.

- Latent class mixed model (LCMM), as implemented in the `lcmm` package.

For the second component, at present the `landmaRk` package supports Cox
proportional hazard models as implemented in the `survival`package. 

Additionally, the `landmaRk` package provides a modular system allowing making 
it possible to incorporate additional models both for the longitudinal and the 
survival components



```{r pipeline-diagram, echo = FALSE, out.width = "100%", fig.cap="Diagram of the landmaRk package pipeline"}
knitr::include_graphics(system.file("diagram.svg", package = "landmaRk"))
```


# Setup

In addition to the `landmaRk` package, we will also use `tidyverse`. 
```{r setup}
set.seed(123)
library(landmaRk)
library(tidyverse)
```

# Example: epileptic data
In this vignette, we use the dataset epileptic to perform landmarking analysis 
of time-to-event data with time-varying covariates. Here is the structure of 
the dataset.

```{r}
data("epileptic")
str(epileptic)
```

The dataset contains the following variables:

-   id: a unique patient identifier

-   time: time when time-varying covariate "dose" was recorded

-   with.time: time when the first of event or censoring happened

-   with.status: indicates whether event (1) or censoring (0) occurred

-   dose: a time-varying covariate

-   treat, age, gender, learn.dis: static (baseline) covariates

# Initialising the landmarking analysis

First, we split the dataset into two, one containing static 
covariates, event time and indicator of event/censoring, and another one 
containing dynamic covariates. To that end, we use the function split_wide_df.
That function returns a named list with the following elements:

- Under the name df_static, a dataframe containing static  covariates, event 
time and indicator of event/censoring.

- Under the name df_dynamic, a named list of dataframes, mapping dynamic 
covariates to dataframes in long format containing longitudinal measurement of 
the relevant dynamic covariate.


```{r}
# DF with Static covariates
epileptic_dfs <- split_wide_df(
  epileptic,
  ids = "id",
  times = "time",
  static = c("with.time", "with.status", "treat", "age", "gender", "learn.dis"),
  dynamic = c("dose"),
  measurement_name = "value"
)
static <- epileptic_dfs$df_static
head(static)
```

```{r}
# DF with Dynamic covariates
dynamic <- epileptic_dfs$df_dynamic
head(dynamic[["dose"]])
```


We can now create an object of class `LandmarkAnalysis`, using the helper 
function of the same name.

```{r}
landmarking_object <- LandmarkAnalysis(
  data_static = static,
  data_dynamic = dynamic,
  event_indicator = "with.status",
  ids = "id",
  event_time = "with.time",
  times = "time",
  measurements = "value"
)
```

Arguments to the helper function are the following:

-   data_static and data_dynamic: the two datasets that were just created.

-   event_indicator: name of the column that indicates the censoring indicator 
in the static dataset.

-   dynamic_covariates: array column names in the dynamic dataset indicating 
time-varying covariates.

-   ids: name of the column that identifies patients in both datasets.

-   event_time: name of the column that identifies time of event/censoring 
in the static dataset.

# Baseline survival analysis (without time-varying covariates)

First, we perform a survival without time-varying covariates. We can use this 
as a baseline to evaluate the performance of a subsequent landmark analysis 
with such covariates. First step is to establish the landmark times, and to 
work out the risk sets at each of those landmark times.

```{r}
landmarking_object <- landmarking_object |>
  compute_risk_sets(landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25))

landmarking_object
```

Now we use the function `fit_survival` to fit the survival model. We specify 
the following arguments:

- `landmarks`: Vector of landmark times at which the model will be fitted.

- `formula`: Two-sided formula object specifying the survival process. Because
we are fitting the baseline model, we include static covariates only.

- `horizons`: Vector of time horizons up to when the model will be fitted. The 
vectors of landmarks and time horizons must be of the same length.
- `method`: Method for the survival component of the landmarking analysis. In 
this case, `"coxph"`.

- `dynamic_covariates`: Vector of names of the dynamic covariates to be used. 
In this baseline analysis, the vector is empty. 

Then, we can make predictions using the function `predict_survival`. Arguments
`method`, `landmarks` and `horizons`are as above. 

```{r}
landmarking_object <- landmarking_object |>
  fit_survival(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    formula = Surv(event_time, event_status) ~ treat + age + gender + learn.dis,
    horizons = seq(from = 2 * 365.25, to = 6 * 365.25, by = 365.25),
    method = "coxph",
    dynamic_covariates = c()
  ) |>
  predict_survival(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    horizons = seq(from = 2 * 365.25, to = 6 * 365.25, by = 365.25),
    method = "coxph",
    type = "survival"
  )
```

To display the results, one can use the method `summary`, specifying 
`type = "survival"`, in addition to a landmark and a horizon. 
```{r}
summary(landmarking_object, type = "survival", landmark = 365.25, horizon = 730.5)
```

Now the `performance_metrics` function can be used to calculate 
(for now, in-sample) performance metrics. 
```{r}
performance_metrics(
  landmarking_object,
  landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
  horizons = seq(from = 2 * 365.25, to = 6 * 365.25, by = 365.25)
)
```

# Landmarking analysis with lme4 + coxph

Now we use the package lme4 to fit a linear mixed model of the time-varying 
covariate, dose. This first step is followed by fitting a Cox PH sub-model 
using the longitudinal predictions as covariates.

```{r}
landmarking_object <- LandmarkAnalysis(
  data_static = static,
  data_dynamic = dynamic,
  event_indicator = "with.status",
  ids = "id",
  event_time = "with.time",
  times = "time",
  measurements = "value"
)

landmarking_object <- landmarking_object |>
  compute_risk_sets(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25)
  )
```

Provided with the risk sets, now the pipeline has the following four steps:

- Fit the longitudinal model with `fit_longitudinal`,
specifying `method = "lme4"`, and the `formula`
as it would be passed to `lme4::lmer()`. One also needs to specify a vector of
`landmarks` and a vector of dynamic covariates, `dynamic_covariates = c("dose")`.

- Make predictions with `predict_longitudinal`, specifying `method = "lme4"`.

- Fit the survival submodel with `fit_survival`, like in the baseline model section,
but specifying the vector of dynamic covariates, `dynamic_covariates = c("dose")`.

- Make predictions with the survival submodel, like in the baseline model section.

```{r}
landmarking_object <- landmarking_object |>
  fit_longitudinal(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    method = "lme4",
    formula = value ~ treat + age + gender + learn.dis + (time | id),
    dynamic_covariates = c("dose")
  ) |>
  predict_longitudinal(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    method = "lme4",
    allow.new.levels = TRUE,
    dynamic_covariates = c("dose")
  ) |>
  fit_survival(
    formula = Surv(event_time, event_status) ~
      treat + age + gender + learn.dis + dose,
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    horizons = seq(from = 2 * 365.25, to = 6 * 365.25, by = 365.25),
    method = "coxph",
    dynamic_covariates = c("dose")
  ) |>
  predict_survival(
    landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
    horizons = seq(from = 2 * 365.25, to = 6 * 365.25, by = 365.25),
    method = "coxph",
    type = "survival"
  )
```

As before, one can also use the function `summary` to display the results.

```{r}
summary(landmarking_object, type = "longitudinal", landmark = 365.25, dynamic_covariate = "dose")
```

```{r}
summary(landmarking_object, type = "survival", landmark = 365.25, horizon = 730.5)
```

Here are the performance metrics: 
```{r}
performance_metrics(
  landmarking_object,
  landmarks = seq(from = 365.25, to = 5 * 365.25, by = 365.25),
  horizons = seq(from = 2 * 365.25, to = 6 * 365.25, by = 365.25)
)
```

# Landmarking analysis with lcmm + coxph

Now we use the
[lcmm package](https://cecileproust-lima.github.io/lcmm/index.html) to fit a
latent class mixed model of the time-varying covariate, dose. This first step is
followed by fitting a Cox PH sub-model using the longitudinal predictions as
covariates.

The pipeline is identical to that of the lme4+coxph model. However, this time
one has to specify `method = "lcmm"` when calling `fit_longitudinal`, 
in addition to the arguments that will be passed on to `lcmm::hlme()`, namely

- `formula`: a two-sided formula of the fixed effects.

- `mixture`: a one-sided formula of the class-specific fixed effects.

- `random`: a one-sided formula specifying the random effects.

- `subject`: name of the column specifying the subject IDs. 

- `ng`: the number of clusters in the LCMM model.

```{r}
landmarking_object <- LandmarkAnalysis(
  data_static = static,
  data_dynamic = dynamic,
  event_indicator = "with.status",
  ids = "id",
  event_time = "with.time",
  times = "time",
  measurements = "value"
)
```

```{r}
landmarking_object <- landmarking_object |>
  compute_risk_sets(seq(from = 365.25, to = 4 * 365.25, by = 365.25)) |>
  fit_longitudinal(
    landmarks = seq(from = 365.25, to = 4 * 365.25, by = 365.25),
    method = "lcmm",
    formula = value ~ treat + age + gender + learn.dis + time,
    mixture = ~ treat + age + gender + learn.dis + time,
    random = ~time,
    subject = "id",
    ng = 2,
    dynamic_covariates = c("dose")
  ) |>
  predict_longitudinal(
    landmarks = seq(from = 365.25, to = 4 * 365.25, by = 365.25),
    method = "lcmm",
    subject = "id",
    avg = FALSE,
    include_clusters = TRUE,
    var.time = "time",
    dynamic_covariates = c("dose")
  ) |>
  fit_survival(
    formula = Surv(event_time, event_status) ~
      treat + age + gender + learn.dis + dose,
    landmarks = seq(from = 365.25, to = 4 * 365.25, by = 365.25),
    horizons = seq(from = 2 * 365.25, to = 5 * 365.25, by = 365.25),
    method = "coxph",
    dynamic_covariates = c("dose"),
    include_clusters = TRUE
  ) |>
  predict_survival(
    landmarks = seq(from = 365.25, to = 4 * 365.25, by = 365.25),
    horizons = seq(from = 2 * 365.25, to = 5 * 365.25, by = 365.25),
    method = "coxph",
    dynamic_covariates = c("dose"),
    include_clusters = TRUE,
    type = "survival",
  )
```

```{r}
summary(landmarking_object,
        type = "longitudinal",
        landmark = 365.25,
        dynamic_covariate = "dose")
```

```{r}
summary(landmarking_object,
        type = "survival",
        landmark = 365.25,
        horizon = 730.5)
```

```{r}
performance_metrics(
  landmarking_object,
  landmarks = seq(from = 365.25, to = 4 * 365.25, by = 365.25),
  horizons = seq(from = 2 * 365.25, to = 5 * 365.25, by = 365.25)
)
```
